The cost of school consolidations: impacts of reduced accessibility

### 2024-04-22 paper/Manuscript.qmd
cdh edits to first half of paper

---
title: "The cost of school consolidations: impacts of reduced accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
linenumbers: true
numbersections: true
format: docx
editor: source
execute: 
  freeze: auto
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: McMaster University SEES
    footnote: "Corresponding Author"
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliation: University of Toronto
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca "
    affiliation: McMaster University CE
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: McMaster University SEES
address:
  - code: McMaster University SEES
    address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - code: McMaster University CE
    address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - code: University of Toronto
    address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"  
abstract: | 
  | The reduction of mobility-induced GHG emissions is a top priority for many communities around the world. From a different standpoint, communities are consistently under pressure to increase operational cost-efficiency savings, including the consideration of spatially consolidating public resource provision. This is the case for public schools, where governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. This study quantifies the travel burdens of school consolidation/closure policy through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. Specifically, we quantify how these closures changed the accessibility landscape for the elementary school aged population by using spatial availability, a novel singly-constrained accessibility measure. Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property. Analysis is conducted at the parcel-level, all school-seats being eligible and travel behaviour based on observved home-to-school patterns for motorzied and non-motorized modes. Analysis on system-wide impacts such as potential GHG emission increases and walk minutes decreases are detailed. Overall, spatial availability decreased in the city. Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pay the price. This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications in policy.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---

@@

Communities are by nature in flux: economic and demographic shifts, rural and urban configurations, and political pressures stoke changes in community characteristics. These forces spur the re-organization and consolidation of public facilities in communities around the world where changes to service provision may improve the quality of service for some but not for all [@Christiaanse_2020; @Rosik_Pulawska-Obiedowska_Goliszek_2021]. Oftentimes, top-down consolidation decisions that are driven by operational cost-savings do not taken into consideration the full extent of benefit loss.

@@

In this study we focus on the spatial travel implications of school closure and consolidation policies. The quantification of lost potential active transportation trips and associated carbon emissions, due to school closure, and the unequal impact attributed to the increased reliance on a motorized commute is scarce in the literature. In this way, our contribution is both methodological and empirical. We quantify the accessibility changes before- and after the first wave of school closure/consolidations implemented in Hamilton, a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in Ontario, Canada [@governmentofcanadaProfileTableCensus2022].

In 2013, the City of Hamilton's largest English school board the Hamilton-Wentworth District School Board (HWDSB), serving the majority of students in the city, released their Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013]. To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subject to accommodation strategies, including a "Accommodation Review" whose outcome is possibly closure/consolidation [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014]. These decisions were in part motivated by operational savings based on a projected reduction in student population and under-utilized school capacity [@craggsAxeCouldFall2012]. However, the case of Hamilton and the HWDSB is unlike others in the province: the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014]. Though the Accommodation Review guidelines outline a community-focused process to deliberate a school's future viability to students, the community and the economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], concerned residences felt a fulsome breakdown of costs to repair schools were not made consistently available [@kleinhuisSeriousConcernsSchool2013]. Further, others pointed to flaud incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013]. Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018]. In a backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards (as they no longer have the ability to levee property taxes) [@mackenzieBlueprintFixOntario2018].

In the first wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed. From between 2011 (before the school closures) to 2016, these closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity, respectively, across the city though overall student population and households in the city did not significantly change between 2011 and 2016 ( `r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2011) and `r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2016)).

From the perspective of public money no longer spent as a result of reduced school operational costs, who and in what way does the Hamilton community pay for these savings by means of accessibility and accessibility-related impacts? In this paper, we hypothesize that that this policy negatively impacted the trips-to-school for many students in addition to decreasing potential active-trip opportunities and increasing emissions. Spatial availability, a singly-constrained competitive accessibility measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], is used to quantify the before- and after- accessibility and associated impacts for the city of Hamilton and the elementary school-aged student population as a result of a province-wide school closure and consolidation policy.

This paper contributes the following: how does the (1) spatial availability, (2) loss in walk minutes and (3) associated carbon emissions change for home-to-school trips after a city closes and consolidates schools? Further, this work examines what communities, based on low-income prevalence, experienced the greatest losses as a result of this policy. The methods introduced in this paper can be used by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy.

The remainder of the paper is organized as follows. Section 2 details the data and methods, particularly multimodal spatial availability, a singly-constrained competitive accessibility measure that considers demand for school-seats by students and friction of distance, as well associated data sources for the inputs into spatial availability. Section 3 showcases the results of the impact that the school consolidation policy had on Hamilton's school spatial availability landscape. Section 4 discusses the results and estimates potential walk minutes lost, where trip-related GHG emissions may have increased, and equity implications for the policy. Section 5 concludes the paper with key remarks. Furthermore, to increase the diffusion of this analysis, all work is completed in a R environment and all associated code and data will be open and reproducible within the lead author's [GitHub repository](https://github.com/soukhova).

@@

Our focus is to evaluate the impact that the school consolidations between 2011 and 2016 had on the multimodal school spatial availability and travel-related GHG emission landscape of the city of Hamilton.

To facilitate comparison, two scenarios are generated. The first and second reflect estimated as-is conditions in 2011 and 2016 assuming the respective school configurations, average student population and low-income prevalence of households, residential parcel locations, and estimated modal travel times in both years. The difference between these scenarios are quantified in terms of accessibility, emission and equity implications and discussed in context with the school consolidation policy.

Data is retrieved at different spatial granularity from the following six data sources.

@@

**Firstly**, schools and associated school catchments are retrieve from the school boards. In Hamilton, the majority of student-aged population attends a school in one of the two English public school boards. One board is referred to as *public* (i.e., Hamilton Wentworth District School Board (HWDSB)) in our work and the other is referred to as *public-catholic* (Hamilton Wentworth Catholic District School Board (HWCDSB)). These catchments indicate what residential property is assigned to what school, by default. Families can decide to attend schools out of catchment, in fact 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized Hamilton-based Hamilton school trips are out of catchment according to the regional travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016 [@data_management_group_tts_2018]. The elementary school locations and catchments are provided by the HWDSB and HWCDSB for the 2010-2011 and 2015-2016 academic years and visualised in @fig-Fig1.

Formally, elementary schools are defined as schools that provide instruction to any combination of grades between kindergarten to grade 8 (i.e. typically children aged 5 to 14). As such, elementary includes middle schools that only instruct grades 6 to 8 (*Mid*), primary schools which only instruct kindergarten to grade 5 or 6 (*JrElem*), and all grade elementary schools (*Elem*) which instruct all grades from kindergarten to grade 8.  The maximum number of students that may be enrolled in a school is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*) and is assigned to each school. The current/historic OTGC province-wide has been made publicly unavailable [@ontarioSchoolFacilityInventory2017]. However, for schools that underwent the full Accommodation Review process, the OTGC for a given year can be scraped from the generated publicly available School Information Profiles. A regression model was derived to estimate the OTGC for schools with no OTGC based on 1) the school's footprint (*F*) in units of (`m^2`), 2) school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m), and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*Mid*, *JrElem*, *Elem*). The building footprint from an archived spatial data set [@DMTI_Spatial_2015] and the 2016 footprints were retrieved from OSM [@OpenStreetMap_2021]. The center of the CBD is defined as a significant intersection in Hamilton Downtown Business Improvement Area [@Hamilton_2014]. This point serves as a pseudo-proxy for age of school construction as the oldest buildings are located within the Hamilton Downtown boundary [@Merrall_2021] and the age of schools is unavailable.

@@

## Students, residential locations and low-income prevelance
Secondly, student population and proportion of household low-income after tax (LIM-AT) prevalence is retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]. The census releases population data by age group category and as pertinent to this paper, 5 to 9 year olds and 10 to 14 year olds. As more relevant to this paper, the LIM-AT prevelance for households with children under 18 (there is no LIM-AT prevelance tabulated for households with exclusively elementary aged children) is retrained. LIM-AT is refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017]. These census variables are taken at the finest spatial unit of analysis disseminated by the Canadian census publicly available, the spatial dissemination area (DA). They are retrieved using the {cancensus} R package [@von_Bergmann_Shkolnik_Jacobs_2021]. DAs are designed by Statistics Canada with the aim of population uniformity hence DAs greatly vary in area but represent between approximately 400 and 700 (1st and 3rd quartile) in total population. The population aged 5 to 14 and the proportion of LIM-AT are visualised in the first and last row in @fig-Fig2.

Thirdly, centroids of residential parcels in 2011 and 2016 are used to represent residential locations [@Teranet_2009]. There are 134,340 and 139,467 residential locations in 2011 and 2016 respectively, representing `r cen_data_2011_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` and `r cen_data_2016_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` households (according to the Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]). As the number of children per residential location is unavailable publicly, the average number of 5 to 14 year olds for each DA is divided by the number of residential households in that DA. Then the potential number of children at each residential location is calculated by multiplying this DA rate. As the [@Teranet_2009] dataset is proprietary, the the middle row in @fig-Fig2 visualises an aggregation of the information: the average rate of 5-14 year olds per residential household at the DA level.

@@

The fourth data source is an estimated travel time matrices from all residential locations to all elementary schools for both motorized (car) and non-motorized (walking) travel. The travel times are estimated using the R-package {r5r} [@Pereira2021r5r], a package that interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022]. For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrick [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the Hamilton boundary). Further, walk travel times that are only less than 27 min minutes were retained. This travel time is approximately equivalent to 1.6 km which is the minimum distance that is required from a school within-catchment to qualify for a school bus [@HWDSB_2019]. <!--put in statistics here-->

It is assumed that children can go to any elementary school, but certain schools are more likely to be attended depending on their proximity to their residential location. Based on this assumption, the fifth data source is empirical data regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018]. The TTS is representative travel survey conducted in the Greater Golden Horseshoe Area in Ontario every 5 years. The trip-level travel data extract for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private care passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home to school in 2011 and 2016 respectively. Data is available at the spatial resolution of Traffic Analysis Zones (TAZ). TAZ are spatial units created for the purpose of the TTS; a few DAs typically nest within each TAZ. {r5r} is used to estimated motorized and non-motorized travel from all TAZ centroids to all TAZ centroids and these estimated travel times are tied to the OD trips. The travel times and the intensity of modal home-to-school flows are visualised in @fig-Fig3 along with the boundaries of the TAZ and DAs. As can be seen in @fig-Fig3, not all TAZ capture an elementary school trip by both modes. For this reason, the proportion of modal-split is aggregated at the community level. Hamilton (Ancaster `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).

@@

The sixth and final data source relates to motorized GHG well-to-wheel estimations. It assumed that all non-motorized trips produce no emissions and those that are motorized take the average emission factor of 0.1 kg CO2e per minute travel. This emission factor is assumed base on an average speed of 40 km/h, the 2,380 g $CO_{2}e$ produced when 1 L of gasoline is combusted and an internal combustion engine passenger vehicle fuel efficiency of 7 g/100km as discussed in @soukhovOccupancyGHGEmissions2022.

@@

As indicators of urban structure, Hansen-type measures are informative, but the meaning of 10,000 accessible school-seats is harder to pin down: how many opportunities must any single student have access to? Furthermore, in scenario comparisons, the interpretation of the percentage change in accessible school-seats is unclear. The interpretability of Hansen-type accessibility has been discussed in numerous studies, including recently by @hu_2019_measuring, @kelobonye2020measuring, and in greater depth by @merlin2017competition along with @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024. The interpretation of accessibility is suggested to depend on how many people demand the opportunity, especially for exclusive opportunity-types like schools-seats (i.e., one school-seat is for one student).

>>>>

---
title: "The cost of school consolidations: impacts of reduced accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
linenumbers: true
numbersections: true
format: docx
editor: source
execute: 
  freeze: auto
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: McMaster University SEES
    footnote: "Corresponding Author"
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliation: University of Toronto
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca "
    affiliation: McMaster University CE
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: McMaster University SEES
address:
  - code: McMaster University SEES
    address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - code: McMaster University CE
    address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - code: University of Toronto
    address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"  
abstract: | 
  | The reduction of mobility-induced GHG emissions is a top priority for many communities around the world. From a different standpoint, communities are consistently under pressure to increase operational cost-efficiency savings, including the consideration of spatially consolidating public resource provision. This is the case for public schools, where governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. This study quantifies the travel burdens of school consolidation/closure policy through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. Specifically, we quantify how these closures changed the accessibility landscape for the elementary school aged population by using spatial availability, a novel singly-constrained accessibility measure. Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property. The analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorzied and non-motorized modes. System-wide impacts resulting from the closures, such as potential GHG emission increases and walk minutes decreases, are detailed. Overall, spatial availability decreased in the city. Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pay the price. This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications associated with policy changes.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---

@@

Communities are by nature in flux: economic and demographic shifts, rural and urban configurations, and political pressures stoke changes in community characteristics. These forces spur the re-organization and consolidation of public facilities in communities around the world where changes to service provision may improve the quality of service for some but not for all [@Christiaanse_2020; @Rosik_Pulawska-Obiedowska_Goliszek_2021]. Oftentimes, top-down consolidation decisions that are driven by operational cost-savings do not take into consideration the full extent of benefit loss.

@@

In this study, we focus on the spatial travel implications of school closure and consolidation policies. The quantification of lost potential active transportation trips and associated carbon emissions due to school closures and the unequal impacts attributed to the increased reliance on a motorized commute is scarce in the literature. In this way, our contribution is both methodological and empirical. We quantify the accessibility changes before- and after the first wave of school closure/consolidations implemented in Hamilton, a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in Ontario, Canada [@governmentofcanadaProfileTableCensus2022].

In 2013, the City of Hamilton's largest English school board the Hamilton-Wentworth District School Board (HWDSB), which is the largest English school board in the city, released their Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013]. To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subject to accommodation strategies, including a "Accommodation Review" whose outcome is possibly closure/consolidation [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014]. These decisions were in part motivated by operational savings based on projected reductions in student population and under-utilized school capacity [@craggsAxeCouldFall2012]. However, the case of Hamilton and the HWDSB is unlike others in the province: the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014]. Though the Accommodation Review guidelines outline a community-focused process to deliberate a school's future viability to students, the community and the economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], concerned residences felt a fulsome breakdown of costs to repair schools were not made consistently available [@kleinhuisSeriousConcernsSchool2013]. Further, others pointed to flawed incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013]. Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018]. In a backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards (as they no longer have the ability to levee property taxes) [@mackenzieBlueprintFixOntario2018].

In the first wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed. From between 2011 (before the school closures) to 2016, these closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity, respectively, across the city though overall student population in the city did not significantly change between 2011 and 2016 ( `r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2011) and `r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2016)).

While accommodation reviews were undertaken to reduce operational costs, how might the closure of schools impact local communities? In this paper, we hypothesize that the school consolidation policy reduced accessibility to schools for many students. By extension, this loss in access resulted in a displacement of trips to school made by active modes that increased transportation-related emissions. To test these hypotheses, the paper examines the following: how does the (1) spatial availability, (2) loss in walk minutes and (3) associated carbon emissions change for home-to-school trips after a city closes and consolidates schools? Further, this work examines (4) what communities, based on low-income prevalence, experienced the greatest losses as a result of this policy. Spatial availability, a singly-constrained competitive accessibility measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], is used to quantify the before- and after- accessibility and associated impacts for the city of Hamilton and the elementary school-aged student population as a result of the school closure and consolidation policy. The methods introduced in this paper can be used by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy. Furthermore, to increase the diffusion of this analysis, all work is completed in a R environment and all associated code and data will be open and reproducible within the lead author's [GitHub repository](https://github.com/soukhova)

 >CUT? The remainder of the paper is organized as follows. Section 2 details the data and methods and Section 3 showcases the results of the impact that the school consolidation policy had on Hamilton's school spatial availability landscape. Section 4 discusses the results and estimates potential walk minutes lost, where trip-related GHG emissions may have increased, and equity implications for the policy. Section 5 concludes the paper with key remarks.

@@

Our focus is to evaluate the impact that the school consolidations between 2011 and 2016 had on the multimodal school spatial availability and travel-related GHG emission landscape of the city of Hamilton. To facilitate comparison, two scenarios are generated. The first and second reflect estimated as-is conditions in 2011 and 2016 assuming the respective school configurations, average student population and low-income prevalence of households, residential parcel locations, and estimated modal travel times in both years. The difference between these scenarios are quantified in terms of accessibility, emission and equity implications and discussed in context with the school consolidation policy. Data is retrieved at different spatial granularity from the following six data sources.

@@

First, schools and associated school catchments are retrieved from the City's school boards. In Hamilton, the majority of student-aged population attends a school in one of the two English public school boards. One board is referred to as *public* (i.e., Hamilton Wentworth District School Board (HWDSB)) in our work and the other is referred to as *public-catholic* (Hamilton Wentworth Catholic District School Board (HWCDSB)). These catchments indicate what residential property is assigned to what school, by default. Families can decide to attend schools out of catchment, in fact 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized school trips are out of catchment according to the regional travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016 [@data_management_group_tts_2018]. The elementary school locations and catchments are provided by the HWDSB and HWCDSB for the 2010-2011 and 2015-2016 academic years and visualised in @fig-Fig1.

Formally, elementary schools are defined as schools that provide instruction to any combination of grades between kindergarten to grade 8 (i.e. typically children aged 5 to 14). As such, elementary includes middle schools that only instruct grades 6 to 8 (*Mid*), primary schools that only instruct kindergarten to grade 5 or 6 (*JrElem*), and all grade elementary schools (*Elem*) which instruct all grades from kindergarten to grade 8. The maximum number of students that may be enrolled in a school is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*) and is assigned to each school. The current/historic OTGC province-wide is not publicly available [@ontarioSchoolFacilityInventory2017]. However, for schools that underwent the full Accommodation Review process, the OTGC for a given year can be obtained from publicly available School Information Profiles. A regression model was derived to estimate the OTGC for schools with no capacity information based on 1) the school's building footprint (*F*) in units of ($m^2$), 2) the school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m), and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*Mid*, *JrElem*, *Elem*). The building footprint from an archived spatial data set [@DMTI_Spatial_2015] and the 2016 footprints were retrieved from OSM [@OpenStreetMap_2021]. Because the age of construction for different schools is not available, we use each school's distance from the centroid of the Downtown Business Improvement Area [@Hamilton_2014] as a proxy for school age as the oldest buildings are generally located closer to the city centre [@Merrall_2021].

@@

## Students, residential locations and low-income prevelance
Second, student population and proportion of household low-income after tax (LIM-AT) prevalence is retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]. The census releases population data by age group category in 5-year increments. The LIM-AT prevelance for households with children under 18 (there is no LIM-AT prevelance tabulated for households with exclusively elementary aged children) is retrained. LIM-AT is refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017]. These census variables are taken at the Dissemination Area (DA) level, which is the finest geographic unit for which census data is made publicly available. Census data is retrieved using the {cancensus} R package [@von_Bergmann_Shkolnik_Jacobs_2021]. DAs are designed by Statistics Canada with the aim of population uniformity hence DAs greatly vary in area but represent between approximately 400 and 700 (1st and 3rd quartile) in total population. The population aged 5 to 14 and the proportion of LIM-AT are visualised in the first and last row in @fig-Fig2.

The third data source consists of centroids of residential parcels in 2011 and 2016 that are used to represent residential locations [@Teranet_2009]. There are 134,340 and 139,467 residential locations in 2011 and 2016 respectively, representing `r cen_data_2011_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` and `r cen_data_2016_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` households (according to the Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]). As the number of children per residential location is unavailable publicly, the average number of 5 to 14 year olds for each DA is divided by the number of residential households in that DA. Then the potential number of children at each residential location is calculated by multiplying this DA rate. Due to the proprietary nature of the parcel data, the middle row in @fig-Fig2 visualises an aggregation of the information: the average rate of 5-14 year olds per residential household at the DA level.

@@

It is assumed that children can go to any elementary school, but certain schools are more likely to be attended depending on their proximity to their residential location. Based on this assumption, the fourth data source is empirical data regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018]. The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario every 5 years. With a target 5% sampling rate, the survey is expanded to be representative at the traffic analysis zone (TAZ) level of geography. TAZ are spatial units created for the purpose of the TTS; a few DAs typically nest within each TAZ. The trip-level travel data extract for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private care passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home to school in 2011 and 2016 respectively. {r5r} is used to estimated motorized and non-motorized travel from all TAZ centroids to all TAZ centroids and these estimated travel times are tied to the OD trips. The travel times and the intensity of modal home-to-school flows are visualised in @fig-Fig3 along with the boundaries of the TAZ and DAs. As can be seen in @fig-Fig3, not all TAZs capture an elementary school trip by both modes. For this reason, the proportion of modal-split is aggregated at the community level. Hamilton (Ancaster `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).

@@

The sixth and final data source relates to motorized GHG well-to-wheel estimations. It assumed that all non-motorized trips produce no emissions and those that are motorized take the average emission factor of 0.1 kg CO2e per minute travel. This emission factor is assumed based on an average speed of 40 km/h, the 2,380 g $CO_{2}e$ produced when 1 L of gasoline is combusted, and an internal combustion engine passenger vehicle fuel efficiency of 7 g/100km as discussed in @soukhovOccupancyGHGEmissions2022.

@@

As indicators of urban structure, Hansen-type measures are informative, but the meaning of 10,000 accessible school-seats is harder to pin down: how many opportunities must any single student have access to? Furthermore, in scenario comparisons, the interpretation of the percentage change in accessible school seats is unclear. The interpretability of Hansen-type accessibility has been discussed in numerous studies, including recently by @hu_2019_measuring, @kelobonye2020measuring, and in greater depth by @merlin2017competition along with @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024. The interpretation of accessibility is suggested to depend on how many people demand the opportunity, especially for exclusive opportunity-types like schools-seats (i.e., one school-seat is for one student).

In this paper, we benefit from new developments in accessibility research, particularly the multimodal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024]. Spatial availability is a singly-constrained accessibility measure that accounts for competition by students using different modes for exclusive opportunities, such as school seats. The measure’s single constraint ensures that the marginals at the destination are met and thus the number of estimated school seats (opportunities) are preserved and allocated proportionally to the mode-using student population. This proportional allocation of opportunities yields an interpretable and meaningful measure of opportunity access, particularly when comparing across modes, at the spatial resolution of a residential parcel, and multiple time periods. See @soukhovMultimodalSpatialAvailability2024 for further discussion, multimodal spatial availability  $V_{i}^m$ is defined as given by Equation (\ref{eq:spatial-availability}).
